Validation of marker gene expression

### 2025-02-20 analyses/cell-type-consensus/exploratory-notebooks/05-marker-gene-validation.Rmd

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(ggplot2)
})

@@

@@

```{r}
# source summarize_celltypes() function
setup_functions <- file.path(module_base, "exploratory-notebooks", "utils", "setup-functions.R")
source(setup_functions)
```

@@

# read in validation markers 
markers_df <- readr::read_tsv(validation_markers_file) |> 
  dplyr::mutate(
    plot_label = glue::glue("{gene_symbol}-{validation_group_annotation}")
  )

validation_groups_df <- readr::read_tsv(validation_groups_file) |> 
  # rename final assigned group to avoid conflicts when merging in marker gene expression 
  # we want to separate the marker gene group from the actual cell type annotation
  dplyr::select(consensus_annotation, assigned_group = validation_group_annotation)
```

```{r}
# combine into one data frame 
combined_df <- consensus_df |> 
  dplyr::left_join(validation_groups_df, by = c("consensus_annotation")) |> 
  dplyr::left_join(gene_exp_df, by = c("barcodes", "library_id")) |> 
  # account for the same gene being present in multiple cell types 
  dplyr::left_join(markers_df, by = c("ensembl_gene_id"), relationship = "many-to-many") |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    detected = logcounts > 0 # column indicating if gene is present or not
  )

# make a data frame that just has the unique genes 
unique_gene_df <- combined_df |> 
  dplyr::filter(gene_observed_count == 1)
```

```{r}
# get total number of cells per final annotation group 
total_cells_df <- unique_gene_df |> 
  dplyr::select(library_id, barcodes, assigned_group) |> 
  unique() |> 
  dplyr::count(assigned_group, name = "total_cells")

# get total number of cells each gene is detected in per group 
# and mean gene expression per group 
group_stats_df <- unique_gene_df |> 
  dplyr::group_by(assigned_group, ensembl_gene_id) |>
  dplyr::summarize(
    detected_count = sum(detected),
    mean_exp = mean(logcounts)
  )

gene_summary_df <- group_stats_df |> 
  # add total cells
  dplyr::left_join(total_cells_df, by = c("assigned_group")) |> 
  dplyr::filter(total_cells > 50) |> 
  # add marker gene groups
  dplyr::left_join(markers_df, by = c("ensembl_gene_id"), relationship = "many-to-many") |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    # get total percent
    percent_exp = (detected_count/total_cells) * 100,
    # account for NA/unknowns and set axes order
    assigned_group = dplyr::if_else(is.na(assigned_group), "unknown", assigned_group) |> 
      factor(levels =  c(unique(markers_df$validation_group_annotation), "unknown"))
  ) 
```

```{r}
# filter out low expressed genes
dotplot_df <- gene_summary_df |> 
  dplyr::filter(mean_exp > 0, percent_exp > 5) |>
  dplyr::arrange(assigned_group) |> 
  dplyr::mutate(y_label = as.factor(glue::glue("{assigned_group} ({total_cells})")))

# pull out cell types to specify legend and x-axis order
celltype_groups <- dotplot_df |> 
  dplyr::pull(assigned_group) |> 
  unique() |>
  as.character()

# specify x axis order
marker_gene_order <- markers_df |> 
  dplyr::filter(validation_group_annotation %in% celltype_groups,
                gene_symbol %in% dotplot_df$gene_symbol) |> 
  dplyr::pull(gene_symbol)

# remove marker genes that aren't present in final annotations and set x axis order 
dotplot_df <- dotplot_df |> 
  dplyr::filter(validation_group_annotation %in% celltype_groups) |> 
  dplyr::mutate(
    gene_symbol = factor(gene_symbol, levels = marker_gene_order)
  )

dotplot <- ggplot(dotplot_df, aes(y = forcats::fct_rev(y_label), x = gene_symbol, color = mean_exp, size = percent_exp)) +
  geom_point() +
  scale_color_viridis_c(option = "magma") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  ) +
  labs(
    x = "",
    y = "Cell type annotation"
  )

color_bar <- ggplot(dotplot_df, aes(x = gene_symbol, y = 1, fill = validation_group_annotation)) + 
  geom_tile() + 
  scale_fill_brewer(palette = "Dark2", breaks = celltype_groups) +
  #scale_fill_manual(values = color_palette, breaks = celltype_groups) +
  ggmap::theme_nothing() +
  theme(legend.position = "bottom") +
  labs(fill = "")

dotplot + color_bar +
  patchwork::plot_layout(ncol = 1, heights = c(4, 0.1))

@@

```{r}
# note that we can't use the same heatmap function as before since we are looking at cell types as rows rather than gene sets
# we also want to specify the annotation name

# get annotation colors for each of the marker gene types 
filtered_markers_df <- markers_df |> 
  dplyr::filter(validation_group_annotation %in% celltype_groups,
                gene_symbol %in% gene_summary_df$gene_symbol)

# first make a mtx to use for the heatmap with rows as cell types and genes as columns 
heatmap_mtx <- gene_summary_df |> 
  dplyr::select(gene_symbol, assigned_group, mean_exp) |> 
  dplyr::filter(gene_symbol %in% filtered_markers_df$gene_symbol) |> 
  tidyr::pivot_wider(
    names_from = gene_symbol,
    values_from = mean_exp
  ) |> 
  tibble::column_to_rownames("assigned_group") |>
  as.matrix()
# make sure cell types are present in the right order
heatmap_mtx <- heatmap_mtx[celltype_groups, filtered_markers_df$gene_symbol]

marker_gene_categories <- unique(filtered_markers_df$validation_group_annotation)
num_categories <- length(marker_gene_categories)
category_colors <- palette.colors(palette = "Dark2") |>
  head(n = num_categories) |>
  purrr::set_names(marker_gene_categories)

# create annotation for heatmap
annotation <- ComplexHeatmap::columnAnnotation(
  category = filtered_markers_df$validation_group_annotation,
  col = list(
    category = category_colors
  ),
  annotation_legend_param = list(
      title = "Marker gene category"
    )
)

ComplexHeatmap::Heatmap(
    heatmap_mtx,
    # set the color scale based on min and max values
    col = circlize::colorRamp2(seq(min(heatmap_mtx), max(heatmap_mtx), length = 2), colors = c("white", "#00274C")),
    border = TRUE,
    ## Row parameters
    cluster_rows = FALSE,
    row_title = "",
    row_title_side = "left",
    row_names_side = "left",
    row_dend_side = "right",
    row_names_gp = grid::gpar(fontsize = 10),
    ## Column parameters
    cluster_columns = FALSE,
    show_column_names = TRUE,
    column_names_gp = grid::gpar(fontsize = 8),
    top_annotation = annotation,
    heatmap_legend_param = list(
      title = "Mean expression"
    )
  )
```

Repeat the plots above but use mean expression of all marker genes across all cells

```{r}
# get total number of cells each gene is detected in per group 
# and mean gene expression per group 
group_stats_df <- combined_df |> 
  dplyr::group_by(assigned_group, ensembl_gene_id) |>
  dplyr::summarize(
    detected_count = sum(detected),
    mean_exp = mean(logcounts),
    .groups = "drop"
  ) |> 
  # add in validation group for marker genes
  dplyr::left_join(markers_df, by = c("ensembl_gene_id")) |>
  dplyr::group_by(assigned_group, validation_group_annotation) |> 
  dplyr::mutate(
    group_mean_exp = mean(mean_exp),
    group_detected_count = mean(detected_count),
    .groups = "drop"
  ) |> 
  dplyr::select(assigned_group, validation_group_annotation, group_mean_exp, group_detected_count) |>
  unique()

gene_summary_df <- group_stats_df |> 
  # add total cells
  dplyr::left_join(total_cells_df, by = c("assigned_group")) |> 
  dplyr::filter(total_cells > 50) |> 
  # add per gene stats
  #dplyr::left_join(group_stats_df, by = c("ensembl_gene_id", "assigned_group")) |> 
  #dplyr::left_join(markers_df, by = c("ensembl_gene_id"), relationship = "many-to-many") |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    # get total percent
    percent_exp = (group_detected_count/total_cells) * 100,
    # account for NA/unknowns and set axes order
    assigned_group = dplyr::if_else(is.na(assigned_group), "unknown", assigned_group) |> 
      factor(levels =  c(unique(markers_df$validation_group_annotation), "unknown"))
  ) 
```

```{r}
# filter out low expressed genes
dotplot_df <- gene_summary_df |> 
  dplyr::filter(group_mean_exp > 0, percent_exp > 5) |>
  dplyr::arrange(assigned_group) |> 
  dplyr::mutate(y_label = as.factor(glue::glue("{assigned_group} ({total_cells})")))

# pull out cell types to specify legend and x-axis order
celltype_groups <- dotplot_df |> 
  dplyr::pull(assigned_group) |> 
  unique() |>
  as.character()

# remove marker genes that aren't present in final annotations and set x axis order 
dotplot_df <- dotplot_df |> 
  dplyr::filter(validation_group_annotation %in% celltype_groups) |>
  dplyr::mutate(
    validation_group_annotation = factor(validation_group_annotation, levels = celltype_groups)
  )

dotplot <- ggplot(dotplot_df, aes(y = forcats::fct_rev(y_label), x = validation_group_annotation, color = group_mean_exp, size = percent_exp)) +
  geom_point() +
  scale_color_viridis_c(option = "magma") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  ) +
  labs(
    x = "Marker gene group",
    y = "Cell type annotation"
  )

color_bar <- ggplot(dotplot_df, aes(x = validation_group_annotation, y = 1, fill = validation_group_annotation)) + 
  geom_tile() + 
  scale_fill_brewer(palette = "Dark2", breaks = celltype_groups) +
  #scale_fill_manual(values = color_palette, breaks = celltype_groups) +
  ggmap::theme_nothing() +
  theme(legend.position = "bottom") +
  labs(fill = "")

dotplot + color_bar +
  patchwork::plot_layout(ncol = 1, heights = c(4, 0.1))

```{r}
# get annotation colors for each of the marker gene types 
filtered_markers_df <- markers_df |> 
  dplyr::filter(validation_group_annotation %in% celltype_groups) |> 
  dplyr::select(validation_group_annotation) |> 
  unique()

# first make a mtx to use for the heatmap with rows as cell types and genes as columns 
heatmap_mtx <- gene_summary_df |> 
  dplyr::select(validation_group_annotation, assigned_group, group_mean_exp) |> 
  tidyr::pivot_wider(
    names_from = validation_group_annotation,
    values_from = group_mean_exp
  ) |> 
  tibble::column_to_rownames("assigned_group") |>
  as.matrix()
# make sure cell types are present in the right order
heatmap_mtx <- heatmap_mtx[celltype_groups, filtered_markers_df$validation_group_annotation]

marker_gene_categories <- unique(filtered_markers_df$validation_group_annotation)
num_categories <- length(marker_gene_categories)
category_colors <- palette.colors(palette = "Dark2") |>
  head(n = num_categories) |>
  purrr::set_names(marker_gene_categories)

# create annotation for heatmap
annotation <- ComplexHeatmap::columnAnnotation(
  category = filtered_markers_df$validation_group_annotation,
  col = list(
    category = category_colors
  ),
  annotation_legend_param = list(
      title = "Marker gene category"
    )
)

ComplexHeatmap::Heatmap(
    heatmap_mtx,
    # set the color scale based on min and max values
    col = circlize::colorRamp2(seq(min(heatmap_mtx), max(heatmap_mtx), length = 2), colors = c("white", "#00274C")),
    border = TRUE,
    ## Row parameters
    cluster_rows = FALSE,
    row_title = "",
    row_title_side = "left",
    row_names_side = "left",
    row_dend_side = "right",
    row_names_gp = grid::gpar(fontsize = 10),
    ## Column parameters
    cluster_columns = FALSE,
    show_column_names = TRUE,
    column_names_gp = grid::gpar(fontsize = 8),
    top_annotation = annotation,
    heatmap_legend_param = list(
      title = "Mean expression"
    )
  )
```

>>>>

This notebook looks at marker gene expression for consensus cell types in `r params$project_id`. 
The marker genes shown in this notebook can be found in [`references/validation-markers.tsv`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-consensus/references/validation-markers.tsv).

All plots will also show the final cell type annotation group based on the `validation_group_annotation` in [`references/consensus-validation-groups.tsv`](https://github.com/AlexsLemonade/OpenScPCA-analysis/blob/main/analyses/cell-type-consensus/references/consensus-validation-groups.tsv). 
This means that cells assigned to more granular cell types will be lumped together with a broader cell type, e.g., memory T cell and naive T cell will be renamed as just T cell.

```{r packages}
suppressPackageStartupMessages({
  # load required packages
  library(ggplot2)
  library(patchwork)
})

@@

```{r}
# marker_heatmap
plotting_functions <- file.path(module_base, "exploratory-notebooks", "utils", "plotting-functions.R")
source(plotting_functions)
```

@@

@@

# read in validation markers 
markers_df <- readr::read_tsv(validation_markers_file)

validation_groups_df <- readr::read_tsv(validation_groups_file) |> 
  # rename final assigned group to avoid conflicts when merging in marker gene expression 
  # we want to separate the marker gene group from the actual cell type annotation
  dplyr::select(consensus_annotation, broad_celltype_group = validation_group_annotation)
```

```{r}
# combine into one data frame that has all marker genes
combined_df <- consensus_df |> 
  dplyr::left_join(validation_groups_df, by = c("consensus_annotation")) |> 
  dplyr::left_join(gene_exp_df, by = c("barcodes", "library_id")) |> 
  # account for the same gene being present in multiple cell types 
  dplyr::left_join(markers_df, by = c("ensembl_gene_id"), relationship = "many-to-many") |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    detected = logcounts > 0 # column indicating if gene is present or not
  ) 
```

## Plot setup

```{r}
# prep for plots 
# get total number of cells per final annotation group 
total_cells_df <- combined_df |> 
  dplyr::select(library_id, barcodes, broad_celltype_group) |> 
  unique() |> 
  dplyr::count(broad_celltype_group, name = "total_cells") 
```

```{r}
# get total number of cells each gene is detected in per group 
# and mean gene expression per group of each marker gene and mean expression of all marker genes 
group_stats_df <- combined_df |> 
  dplyr::group_by(broad_celltype_group, ensembl_gene_id) |>
  dplyr::summarize(
    detected_count = sum(detected),
    mean_exp = mean(logcounts)
  ) |> 
  # add in validation group for marker genes
  dplyr::left_join(markers_df, by = c("ensembl_gene_id"), relationship = "many-to-many") |>
  dplyr::group_by(broad_celltype_group, validation_group_annotation) |> 
  dplyr::mutate(
    # calculate mean expression/detected across all markers for a specific group 
    all_markers_mean_exp = mean(mean_exp),
    all_markers_detected_count = mean(detected_count)
  ) |>  # add total cells
  dplyr::left_join(total_cells_df, by = c("broad_celltype_group")) |> 
  # for plotting we're only going to look at any cell types with > 50 cells otherwise these plots can get wild 
  dplyr::filter(total_cells > 50) |> 
  dplyr::rowwise() |> 
  dplyr::mutate(
    # get total percent expressed
    percent_exp = (detected_count/total_cells) * 100,
    all_markers_percent_exp = (all_markers_detected_count/total_cells) * 100, 
    # account for NA/unknowns and set axes order
    broad_celltype_group = dplyr::if_else(is.na(broad_celltype_group), "unknown", broad_celltype_group) |> 
      factor(levels =  c(unique(markers_df$validation_group_annotation), "unknown"))
  )

# make a data frame that just has the unique genes 
unique_gene_df <- group_stats_df |> 
  dplyr::filter(gene_observed_count == 1)
```

```{r}
# get list of celltypes to keep and assign colors 
celltype_groups <- group_stats_df |> 
  dplyr::pull(broad_celltype_group) |> 
  unique() |>
  as.character()

# assign colors to use in annotation bars 
celltype_colors <- palette.colors(palette = "alphabet")[1:length(celltype_groups)]
names(celltype_colors) <- celltype_groups

# set order for cell types 
celltype_order <- unique(filtered_markers_df$validation_group_annotation)
```

```{r}
# filter markers to those that are actually relevant 
# set order of gene symbol based on celltype groups 
filtered_markers_df <- markers_df |> 
  dplyr::filter(gene_observed_count == 1,
                validation_group_annotation %in% celltype_groups,
                gene_symbol %in% group_stats_df$gene_symbol)

# specify x axis order
marker_gene_order <- markers_df |> 
  dplyr::filter(validation_group_annotation %in% celltype_groups,
                gene_symbol %in% dotplot_df$gene_symbol) |> 
  dplyr::pull(gene_symbol)
```

## Expression of unique marker genes

First we will look at the expression of marker genes that are specific to a cell type group.  
All marker genes shown in the following plots are only found to be a marker gene for a single cell type group. 
Only cell type groups with at least 50 cells will be shown.

Note that I did try and look at all marker genes, without specifying that they must be unique, and the plots were overwhelming with lots of genes. 
Instead I included the a separate section looking at the mean expression across all markers for a given group.

### Dot plot

The dot plot shows the mean expression of each marker gene across all cells assigned to a given cell type group. 
Only genes that have mean expression > 0 and are expressed in at least 5% of cells in at least one cell type group will be shown.

```{r, fig.height=10, fig.width=15}
# filter out low expressed genes
dotplot_df <- unique_gene_df |> 
  dplyr::filter(mean_exp > 0, percent_exp > 5) |> 
  dplyr::arrange(broad_celltype_group) |> 
  # add a label for the plot 
  dplyr::mutate(y_label = as.factor(glue::glue("{broad_celltype_group} ({total_cells})"))) |> 
  # remove marker genes that aren't present in final annotations and set x axis order 
  dplyr::filter(gene_symbol %in% marker_gene_order) |> 
  dplyr::mutate(
    # set orders of gene symbol and validation groups 
    gene_symbol = factor(gene_symbol, levels = marker_gene_order),
    validation_group_annotation = factor(validation_group_annotation, levels = celltype_order)
  )

# make dotplot with marker gene exp
dotplot <- ggplot(dotplot_df, aes(y = forcats::fct_rev(y_label), x = gene_symbol, color = mean_exp, size = percent_exp)) +
  geom_point() +
  scale_color_viridis_c(option = "magma") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  ) +
  labs(
    x = "",
    y = "Broad cell type annotation",
    color = "Mean gene expression",
    size = "Percent cells expressed"
  )

# add annotation bar aligning marker genes with validation group 
color_bar <- ggplot(dotplot_df, aes(x = gene_symbol, y = 1, fill = validation_group_annotation)) + 
  geom_tile() + 
  scale_fill_manual(values = plot_colors, breaks = levels(dotplot_df$validation_group_annotation)) +
  ggmap::theme_nothing() +
  theme(legend.position = "bottom") +
  labs(fill = "")

dotplot + color_bar +
  patchwork::plot_layout(ncol = 1, heights = c(4, 0.1)) 
```

### Heatmap

The heatmap shows the mean expression of each marker gene across all cells assigned to a given cell type group.

```{r}
# first make a mtx to use for the heatmap with rows as cell types and genes as columns 
unique_heatmap_mtx <- unique_gene_df |> 
  dplyr::ungroup() |> 
  dplyr::select(gene_symbol, broad_celltype_group, mean_exp) |> 
  dplyr::filter(gene_symbol %in% filtered_markers_df$gene_symbol) |> 
  tidyr::pivot_wider(
    names_from = gene_symbol,
    values_from = mean_exp
  ) |> 
  tibble::column_to_rownames("broad_celltype_group") |>
  as.matrix()
# make sure cell types are present in the right order
unique_heatmap_mtx <- unique_heatmap_mtx[c(celltype_order, "unknown"), filtered_markers_df$gene_symbol]

# create annotation for heatmap
annotation <- ComplexHeatmap::columnAnnotation(
  group = filtered_markers_df$validation_group_annotation,
  col = list(
    group = celltype_colors
  ),
  annotation_legend_param = list(
      title = "Validation group"
    )
)
```

```{r, fig.height=7, fig.width=10}
marker_gene_heatmap(unique_heatmap_mtx, annotation)
```

## Mean expression across all marker genes

We will make the same plots but instead of looking at the expression of individual marker genes for all cell type groups, we will plot the mean expression of all marker genes for a specific cell type group.

This is calculated by first getting the mean expression of each individual marker gene across all cells in a cell type group and then taking the mean of that value across all marker genes. 
This results in one row per assigned cell type group and one column per marker gene group.

### Dot plot

```{r}
# filter low expressed genes 
grouped_dotplot_df <- group_stats_df |> 
  dplyr::select(broad_celltype_group, validation_group_annotation, all_markers_mean_exp, all_markers_percent_exp, total_cells) |>
  unique() |> 
  dplyr::filter(all_markers_mean_exp > 0, all_markers_percent_exp > 5, validation_group_annotation %in% celltype_groups) |>
  dplyr::arrange(broad_celltype_group) |> 
  # make sure labels are correct 
  dplyr::mutate(y_label = as.factor(glue::glue("{broad_celltype_group} ({total_cells})")),
                validation_group_annotation = factor(validation_group_annotation, levels = celltype_order))

ggplot(grouped_dotplot_df, aes(y = forcats::fct_rev(y_label), x = validation_group_annotation, color = all_markers_mean_exp, size = all_markers_percent_exp)) +
  geom_point() +
  scale_color_viridis_c(option = "magma") +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)
  ) +
  labs(
    x = "Marker gene group",
    y = "Broad cell type annotation", 
    color = "Mean gene expression", 
    size = "Percent cells expressed"
  )

@@

### Heatmap

```{r}
# get annotation 
grouped_annotation <- ComplexHeatmap::columnAnnotation(
  group = unique(filtered_markers_df$validation_group_annotation),
  col = list(
    group = celltype_colors
  ),
  annotation_legend_param = list(
      title = "Validation group"
    )
)

# first make a mtx to use for the heatmap with rows as cell types and marker gene groups as columns 
grouped_heatmap_mtx <- group_stats_df |> 
  dplyr::select(validation_group_annotation, broad_celltype_group, all_markers_mean_exp) |> 
  unique() |>
  dplyr::filter(validation_group_annotation %in% celltype_order) |> 
  tidyr::pivot_wider(
    names_from = validation_group_annotation,
    values_from = all_markers_mean_exp
  ) |> 
  tibble::column_to_rownames("broad_celltype_group") |>
  as.matrix()
```

```{r}
# make sure cell types are present in the right order
grouped_heatmap_mtx <- grouped_heatmap_mtx[c(celltype_order, "unknown"), celltype_order]
```

```{r, fig.height=7, fig.width=10}
marker_gene_heatmap(grouped_heatmap_mtx, grouped_annotation)
```

## Conclusions

- Regardless of looking at individual genes or the mean expression of all genes, we see similar patterns where generally the cells assigned to a given cell type show expression of the appropriate marker genes when compared to other cells. 
- There seems to be pretty clear distinction of T cells, B cells, and NK cells. 
- Monocytes, macrophages, and myeloid cells are generally hard to distinguish from each other and all share similar marker genes. 
Perhaps we could collapse these into myeloid, but I do see expression of monocyte specific genes in monocytes only and maybe some macrophage specific genes in macrophages only, so it's probably okay to keep separate. 
I think that is more obvious in the heatmaps than in the dot plots. 
- Neurons and astrocytes both show expression of both neuron and astrocyte marker genes. 
It does look like there are a set of genes that are more highly expressed in the astrocyte population though, so I think we can keep those gene sets separate. 
- Here the unknown cells don't show strong expression of any specific set of genes, but it does look like they show more expression of neuron/astrocyte markers than any of the other marker genes in the plot. 
This makes sense since we are looking at brain tumors and I expect similarity between tumor cells and neural cell types.
